Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-11-03 paper.tex

An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space. Social agents thus constitute the microscopic level of urban systems, which are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects. Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}. Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}. Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. \\
Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average. On the contrary, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time provide the use of ergodic models. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. \\

>>>>

An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space. Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects. Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framewo emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}. Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}. Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. \\
Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average. On the contrary, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time provide the use of ergodic models. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. \\
